### 2021-07-08 rad-icfp-2021/abstract.tex
.

\begin{abstract}
    Deep learning is moving towards increasingly sophisticated optimization objectives that employ
    tensors and operations on tensors.
    Reverse-mode differentiation is used for optimization, 
    but it does not preserve purity of the underlying program and makes it notoriously harder to optimize.
    In a purely functional setting, most array operations can be expressed conveniently using map and fold.
    We present a differentiable programming language which is the first to deliver on efficient, pure and parallel-friendly reverse-mode differentiation on
    an expressive array language. 
    We show that our transformation is semantically correct and verifies the cheap gradient principle. 
    Our transformation is presented as a compilation scheme through a novel intermediate representation that we call `unary form`.
    We obtain provably efficient gradients by performing general partial evaluation optimizations after our reverse-mode transformation, as opposed to manually derived ones.
    For simple first-order programs, the obtained output programs resemble static single assignment (SSA) code. 
    We emphasize the modularity of our approach and show how our language can easily be enriched with more optimized primitives, as required for some speedups in practice.
\end{abstract}

>>>>

\begin{abstract}
    Reverse-mode differentiation is used for optimization, but it does not preserve purity of the underlying program which makes it notoriously harder to optimize.
    We present a reverse-mode differentiation on a purely functional language with array operations.  
    It is the first one to deliver a provably efficient, purely functional and denotationnally correct reverse-mode differentiation.
    We show that our transformation is semantically correct and verifies the cheap gradient principle. 
    Inspired by PROPs and compilation to categories, we introduce a novel intermediate representation that we call `unary form'.
    Our reverse-mode transformation is factored as a compilation scheme through this intermediate representation.
    We obtain provably efficient gradients by performing general partial evaluation optimizations after our reverse-mode transformation, as opposed to manually derived ones.
    For simple first-order programs, the obtained output programs resemble static single assignment (SSA) code. 
    We emphasize the modularity of our approach and show how our language can easily be enriched with more optimized primitives, as required for some speedups in practice.
\end{abstract}

### 2021-07-08 rad-icfp-2021/background.tex

Let the term we want to differentiate be $x1:\RR,\ldots,xn:\RR\vdash exp(cos(xi))$.
Following the chain rule, we need the jacobian matrices of $cos$ at $xi$ and of $exp$ at $cos(xi)$. 
Instead of considering these as operations from $\RR\to\RR$, we consider them as functions from the whole context. So $cos$ and $exp$ are seen as functions $\RR^n\to\RR$.
However, by simply doing this we lose compositionality. 
So $cos$ will also return its context and is now seen as a function $\sem{cos}:\RR^n\to\RR^{n+1}$.
Similarly for $exp$, but also taking the return value of $\sem{cos}$ as an extra argument, the one it will actually use and not simply return. 
We thus obtain $\sem{exp}:\RR^{n+1}\to\RR^{n+2}$. Now their jacobians matrices $J\sem{cos} \in Mat_{n,n+1}$, $J\sem{exp} \in Mat_{n+1,n+2}$ also compose nicely.
The same can be done for binary (or more) operators and let bindings (assuming no nested lets). 
This transforms a first-order program to a function $f:\RR^n\to\RR^{n+m}$ of the form $f_m\circ\ldots\circ f_1$. 
If the original program was of type $\RR$, then the return value of the original program is the last component of $f$.
Following the mathematical presentation of reverse-mode above, the gradient of the the original program is then obtained as 
\begin{center}
    \begin{tabular}{r c l}
        $\nabla f$ &=& $J^Tf(0,\ldots,0,1)=J^Tf_1(J^Tf_{2}(\ldots(J^Tf_m(0,\ldots,0,1))\ldots)$
    \end{tabular}
\end{center}

@@

\smartpara{Insight 3}
We know in advance that the jacobian functions are going to be applied one to another and we can use partial evaluation to $\beta$-reduce all of these $\lambda$s.
Because each function is almost the identity, we obtain a lot of substitutions of the form $[x/y]$ where both $x$ and $y$ are variables. 
This allows us to drastically reduce the size of $J^Tf$. In fact, for simple programs, this is basically enough to get an efficient purely functional reverse derivative transformation.
We develop this idea further for a richer language.

>>>>

Let the term we want to differentiate be $x_1:\RR,\ldots,x_n:\RR\vdash \exp(\cos(x_i))$.
Following the chain rule, we need the jacobian matrices of $\cos$ at $x_i$ and of $\exp$ at $\cos(x_i)$. 
Instead of considering these as operations from $\RR\to\RR$, we consider them as functions from the whole context. So $\cos$ and $\exp$ are seen as functions $\RR^n\to\RR$.
However, by simply doing this we lose compositionality. 
So $\cos$ will also return its context and is now seen as a function $\sem{\cos}:\RR^n\to\RR^{n+1}$.
Similarly for $\exp$, but also taking the return value of $\sem{\cos}$ as an extra argument, the one it will actually use and not simply return. 
We thus obtain $\sem{\exp}:\RR^{n+1}\to\RR^{n+2}$. Now their jacobians matrices $\J\sem{\cos} \in Mat_{n,n+1}$, $\J\sem{\exp} \in Mat_{n+1,n+2}$ also compose nicely.
The same can be done for binary (or more) operators and let bindings (assuming no nested lets). 
This transforms a first-order program to a function $f:\RR^n\to\RR^{n+m}$ of the form $f_m\circ\ldots\circ f_1$. 
If the original program was of type $\RR$, then the return value of the original program is the last component of $f$.
Following the mathematical presentation of reverse-mode above, the gradient of the the original program is then obtained as 
\begin{center}
    \begin{tabular}{r c l}
        $\nabla f$ &=& $\J^Tf(0,\ldots,0,1)= \J^Tf_1(J^Tf_{2}(\ldots(\J^Tf_m(0,\ldots,0,1))\ldots)$
    \end{tabular}
\end{center}

@@

\smartpara{Insight 3}
We know in advance that the jacobian functions are going to be applied one to another and we can use partial evaluation to $\beta$-reduce all of these $\lambda$s.
Because each function is almost the identity, we obtain a lot of substitutions of the form $[x/y]$ where both $x$ and $y$ are variables. 
This allows us to drastically reduce the size of $\J^Tf$. In fact, for simple programs, this is basically enough to get an efficient purely functional reverse derivative transformation.
We develop this idea further for a richer language.
